How do changes in ex vary between countries?

### 2022-04-11 draft_working_paper.Rmd
add diagnostic statistical plots comparing breakpoints between methods

```{r, echo=FALSE}
# load packages
pacman::p_load(knitr, here, tidyverse, flextable) # using flextable instead of gt. It seems to be better 
# developed, more commonly used, and has more clearly labelled function names.
# load data 
hmd_lt <- read_rds(here("data", "lifetables.rds"))
```

@@

- [ ] make smoother lines coloured by sex
- [X] use consistent colours for all graphs showing separate sex results

@@

- [ ] Use slightly stronger (and consistent) colours for sex smoother lines

@@

### Scotland
```{r, echo = FALSE, warning=FALSE, message = FALSE}
#SCO female at birth
f_e0_SCO <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(x == 0) %>% 
      filter(sex == "female") %>% 
      filter(code == "GBR_SCO") %>% 
      filter(year >= 1979)

@@

```{r, echo=FALSE, tab.cap = "Source: HMD"}
hmd_ex_selected_countries_with_synth %>% 
    filter(year >= 1979) %>% 
    group_by(code, x, sex) %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>%
    mutate(decade = year - (year %% 10)) %>% 
    filter(!is.na(delta_ex)) %>% 
  group_by(code, x, sex, decade) %>% 
  summarise(
    mean_delta_ex = mean(delta_ex),
    sd_delta_ex   = sd(delta_ex)
    ) %>% 
  ungroup() %>% 
  mutate(cell_contents = glue::glue("{round(mean_delta_ex, 3)} ({round(sd_delta_ex, 3)})")
  ) %>% 
  filter(code == "GBRTENW") %>% 
  filter(x == 0) %>% 
  select(sex, decade, cell_contents) %>% 
  spread(decade, cell_contents) %>% 
#  write.csv("clipboard")
  flextable() %>%
  # Want to make the header row bold
  vline(j = 1, part = "all") %>%
  set_caption(caption = "Average improvement in life expectancy by decade")

@@

# Let's see if this now works in the functional programming context, using 
# nest and tibble elements 
hmd_ex_selected_countries_with_synth %>%
  filter(year >= 1979) %>% 
  group_by(code, x, sex) %>% 
  nest() %>% 
  mutate(
    best_tau = map_dbl(data, get_best_tau_from_optim)
  )
# I think this is getting stuck on local optima, so instead maybe 
# the more exhaustive search should be used after all! 
# There are two alternative approaches: 
# 1) Use an exhaustive grid search approach 
# 2) Use an algorithm which is less likely to get stuck in local optima
# The first approach is much more sensible, but the second might be a cool
# stretch goal as well! 
# let's use a heavily modified version of the above function to do the grid search

compare_best_breakmodel_with_null <- 
  hmd_ex_selected_countries_with_synth %>%
    filter(year >= 1979) %>% 
    group_by(code, x, sex) %>% 
    nest() %>% 
    mutate(
      best_tau = map_dbl(data, get_best_tau_from_gridsearch)
    ) %>% 
    mutate(
      model_null = map(
        data, 
        ~ .x %>% 
          arrange(year) %>% 
          mutate(delta_ex = ex - lag(ex)) %>% 
          filter(!is.na(delta_ex)) %>% 
          lm(delta_ex ~ 1, data = .)
      )
    ) %>% 
    mutate(
      model_alt = map2(
        best_tau, data,
      ~ .y %>% 
        arrange(year) %>% 
        mutate(delta_ex = ex - lag(ex)) %>% 
        filter(!is.na(delta_ex)) %>% 
        mutate(T_param = ifelse(year < .x, FALSE, TRUE)) %>% 
        lm(delta_ex ~ T_param, data = .)
      )
    )
tmp <- compare_best_breakmodel_with_null %>% 
  mutate(
    aic_alt  = map_dbl(model_alt, AIC),
    aic_null = map_dbl(model_null, AIC)
  ) %>% 
  mutate(
    alt_preferred = map2_lgl(aic_alt, aic_null, ~ifelse(.x < .y, TRUE, FALSE))
  ) %>% 
  mutate(
    anova_results = map2(model_alt, model_null, anova)
  ) %>% 
  mutate(
    ftest_pval = map_dbl(anova_results, ~.[2,6]),
    stat_sig_preference = ftest_pval < 0.05
  )

@@

@@

@@

@@

- [ ] Jon to figure out why this doesn't work...

@@

```{r}
stat_sig_table <- select(tmp, code, x, sex, best_tau, ftest_pval, stat_sig_preference) %>% arrange(ftest_pval)

@@

>>>>

```{r, echo=FALSE}
# load packages
pacman::p_load(knitr, here, segmented, tidyverse)

# load data 
hmd_lt <- read_rds(here("data", "lifetables.rds"))
```

@@

- [X] make smoother lines coloured by sex
- [X] use consistent colours for all graphs showing separate sex results

@@

@@

### Scotland

```{r, echo = FALSE, warning=FALSE, message = FALSE}
#SCO female at birth
f_e0_SCO <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(x == 0) %>% 
      filter(sex == "female") %>% 
      filter(code == "GBR_SCO") %>% 
      filter(year >= 1979)

@@

```{r, echo=FALSE, tab.cap = "Source: HMD"}
hmd_ex_selected_countries_with_synth %>% 
    filter(year >= 1979) %>% 
    group_by(code, x, sex) %>% 
    arrange(year) %>% 
    mutate(delta_ex = ex - lag(ex)) %>%
    mutate(decade = year - (year %% 10)) %>% 
    filter(!is.na(delta_ex)) %>% 
  group_by(code, x, sex, decade) %>% 
  summarise(
    mean_delta_ex = mean(delta_ex),
    sd_delta_ex   = sd(delta_ex)
    ) %>% 
  ungroup() %>% 
  mutate(cell_contents = glue::glue("{round(mean_delta_ex, 3)} ({round(sd_delta_ex, 3)})")
  ) %>% 
  filter(code == "GBRTENW") %>% 
  filter(x == 0) %>% 
  select(sex, decade, cell_contents) %>% 
  spread(decade, cell_contents) #%>% 
#  write.csv("clipboard")

@@

# Let's see if this now works in the functional programming context, using 
# nest and tibble elements 
# hmd_ex_selected_countries_with_synth %>%
#   filter(year >= 1979) %>% 
#   group_by(code, x, sex) %>% 
#   nest() %>% 
#   mutate(
#     best_tau = map_dbl(data, get_best_tau_from_optim)
#   )
# I think this is getting stuck on local optima, so instead maybe 
# the more exhaustive search should be used after all! 
# There are two alternative approaches: 
# 1) Use an exhaustive grid search approach 
# 2) Use an algorithm which is less likely to get stuck in local optima
# The first approach is much more sensible, but the second might be a cool
# stretch goal as well! 
# let's use a heavily modified version of the above function to do the grid search

compare_best_breakmodel_with_null <- 
  hmd_ex_selected_countries_with_synth %>%
    filter(year >= 1979) %>% 
    group_by(code, x, sex) %>% 
    nest() %>% 
    mutate(
      best_tau = map_dbl(data, get_best_tau_from_gridsearch),
      all_taus = map(data, get_best_tau_from_gridsearch, what = "all")
    ) %>% 
    mutate(
      model_null = map(
        data, 
        ~ .x %>% 
          arrange(year) %>% 
          mutate(delta_ex = ex - lag(ex)) %>% 
          filter(!is.na(delta_ex)) %>% 
          lm(delta_ex ~ 1, data = .)
      )
    ) %>% 
    mutate(
      model_alt = map2(
        best_tau, data,
      ~ .y %>% 
        arrange(year) %>% 
        mutate(delta_ex = ex - lag(ex)) %>% 
        filter(!is.na(delta_ex)) %>% 
        mutate(T_param = ifelse(year < .x, FALSE, TRUE)) %>% 
        lm(delta_ex ~ T_param, data = .)
      )
    ) %>% 
  mutate(
    aic_alt  = map_dbl(model_alt, AIC),
    aic_null = map_dbl(model_null, AIC)
  ) %>% 
  mutate(
    alt_preferred = map2_lgl(aic_alt, aic_null, ~ifelse(.x < .y, TRUE, FALSE))
  ) %>% 
  mutate(
    anova_results = map2(model_alt, model_null, anova)
  ) %>% 
  mutate(
    ftest_pval = map_dbl(anova_results, ~.[2,6]),
    stat_sig_preference = ftest_pval < 0.05
  )

@@

Let's now try to summarise the above results intelligently.

We want to know:

- [ ] Best year
- [ ] Whether stat sig over null 
- [ ] Estimate before break
- [ ] Estimate after break

Let's take the object above then get this info.

```{r}
compare_best_breakmodel_with_null %>% 
  mutate(population = glue::glue("{code}_{sex}_{x}")) %>% 
  ggplot(aes(best_tau, fct_reorder(population, best_tau), colour = stat_sig_preference)) + 
  geom_point()

```
I'm now a bit concerned by how many picked 2015. Is this an artefact of when the data finished?

To test this further I've now saved AIC by proposed year so let's look at how clear a 'winner' is for each population

```{r}
compare_best_breakmodel_with_null %>% 
  select(code, x, sex, best_tau, all_taus, ftest_pval) %>% 
  unnest(all_taus) %>% 
  ungroup() %>% 
  mutate(population = glue::glue("{code}_{x}_{sex}")) %>% 
  mutate(
    ftest_cat = case_when(
      ftest_pval < 0.01 ~ "<0.01",
      between(ftest_pval, 0.01, 0.05) ~ "0.01-0.05",
      between(ftest_pval, 0.05, 0.10) ~ "0.05-0.10",
      TRUE                            ~ "Not Stat. Sig."
    ),
    ftest_cat = factor(ftest_cat, 
      levels = c("<0.01", "0.01-0.05", "0.05-0.10", "Not Stat. Sig."),
      ordered = TRUE
    )
  ) %>% 
  ggplot(aes(tau, aic)) +
  geom_line() + geom_point(aes(colour = ftest_cat, alpha = tau == best_tau)) + 
  facet_wrap(~population, scales = "free_y") +
  scale_colour_manual(
    values = c(
      "<0.01" = "darkred", 
      "0.01-0.05"  = "red", 
      "0.05-0.10" = "orange", 
      "Not Stat. Sig." = "lightblue"
      )
  ) + 
  scale_alpha_manual(values = c(`FALSE` = 0, `TRUE` = 1), guide = "none") +
  labs(
    title = "Penalised fit (AIC) against proposed changepoint year, various population",
    subtitle = "Best fit identified indicate with a point, whose colour indicates statistical significance of changepoint model compared with Null model",
    colour = "Statistical\nsignificance", 
    x = "Year proposed",
    y = "AIC (lower is better)"
  )
```

I think this is a fairly complex visualisation, but hopefully usable either in the main paper or (more likely) supplementary material.

Here's a table attempting to summarise (subjectively) the above graphs:

| **Population**    | **x** | **Sex** | **Main Breakpoint (Stat sig.)** | **Other possible breakpoints** |
|-------------------|-------|---------|---------------------------------|--------------------------------|
| East Germany      | 0     | male    | 1991 **                         | No                             |
| East Germany      | 0     | female  | 2005 **                         | 2015                           |
| East Germany      | 65    | male    | 1991 **                         | No                             |
| East Germany      | 65    | female  | 1991 **                         | No                             |
| Germany           | 0     | male    | 2015 **                         | 2007ish                        |
| Germany           | 0     | female  | 2007 **                         | 2005-2008                      |
| Germany           | 65    | male    | 2015 **                         | No                             |
| Germany           | 65    | female  | 2015 *                          | 2005-2008                      |
| Synthetic Germany | 0     | male    | 2015 **                         | Weak 2007                      |
| Synthetic Germany | 0     | female  | 2007 **                         | 2005-2008, 2015                |
| Synthetic Germany | 65    | male    | 2015 **                         | No                             |
| Synthetic Germany | 65    | female  | 2015 *                          | Weak 2005-2007                 |
| West Germany      | 0     | male    | 2015 **                         | Weak 2007                      |
| West Germany      | 0     | female  | 1988 **                         | 2005-2008ish                   |
| West Germany      | 65    | male    | 2015 **                         | No                             |
| West Germany      | 65    | female  | 2015 *                          | 2008 ish (Weak)                |
| Spain             | 0     | male    | 1983 *                          | No                             |
| Spain             | 0     | female  | 1983 *                          | No                             |
| Spain             | 65    | male    | 2000 (NS)                       | 2015                           |
| Spain             | 65    | female  | 2014 (NS)                       | 1983?                          |
| France            | 0     | male    | 2015 (NS)                       | ?                              |
| France            | 0     | female  | 2015 (NS)                       | ?                              |
| France            | 65    | male    | 2015 **                         | No                             |
| France            | 65    | female  | 2015 *                          | No                             |
| Scotland          | 0     | male    | 2015 ***                        | No                             |
| Scotland          | 0     | female  | 1980 *                          | 2015 (Strong)                  |
| Scotland          | 65    | male    | 1994 (NS)                       | No?                            |
| Scotland          | 65    | female  | 1981 (NS)                       | No                             |
| England & Wales   | 0     | male    | 2013 ***                        | 2012-2015ish                   |
| England & Wales   | 0     | female  | 2012 **                         | Around 2015                    |
| England & Wales   | 65    | male    | 2012 *                          | 2015                           |
| England & Wales   | 65    | female  | 2012 (NS)                       | 2015                           |
| Italy             | 0     | male    | 1981 (NS)                       | 2015                           |
| Italy             | 0     | female  | 2005 (NS)                       | Weak 2015                      |
| Italy             | 65    | male    | 1984 (NS)                       | 1981                           |
| Italy             | 65    | female  | 2005 (NS)                       | 1984                           |

@@

estimate_breakpoint_and_pval_v2 <- function(df){
  null_mdl <- lm(ex ~ year, data = df) 
  seg_mdl <- segmented::segmented(null_mdl, seg.Z= ~year, psi = 2010)   
  anova_diff <- anova(null_mdl, seg_mdl)

segmented_breakpoint_models_v2 <- 
  hmd_ex_selected_countries_with_synth %>% 
    filter(year >= 1979) %>% 
    group_by(code, x, sex) %>% 
    nest() %>% 
    mutate(
      mdl_outputs = map(data, estimate_breakpoint_and_pval_v2)
    ) %>% 
    unnest_wider(mdl_outputs) %>% 
    mutate(
      ftest_pval = map_dbl(diff, ~.[2,6]),
      stat_sig_preference = ftest_pval < 0.05
    ) %>% 
    mutate(
      breakpoint = map_dbl(seg, ~.[["psi"]][1,2]),
      breakpoint_se = map_dbl(seg, ~.[["psi"]][1,3])
    )

segmented_breakpoint_models_v2

# To check the results are the same more formally

identical(
  segmented_breakpoint_models %>% 
    select(
      code, x, sex, ftest_pval, stat_sig_preference, breakpoint
    ), 
  segmented_breakpoint_models_v2 %>% 
    select(
      code, x, sex, ftest_pval, stat_sig_preference, breakpoint
    )
  )

# Returns FALSE

# How are they different? 
models_comparison <- 
segmented_breakpoint_models %>% 
  select(code, x, sex, ftest_pval, stat_sig_preference, breakpoint, breakpoint_se) %>% 
  group_by(code, x, sex) %>% 
  nest(v1 = c(ftest_pval, stat_sig_preference, breakpoint, breakpoint_se)) %>% 
  left_join(
    segmented_breakpoint_models_v2 %>% 
      select(code, x, sex, ftest_pval, stat_sig_preference, breakpoint, breakpoint_se) %>% 
      group_by(code, x, sex) %>% 
      nest(v2 = c(ftest_pval, stat_sig_preference, breakpoint, breakpoint_se))
  ) %>% 
  unnest_wider(
    v1, names_sep = "."
  ) %>% 
  unnest_wider(
    v2, names_sep = "."
  )

# How do the breakpoints and SEs differ? 
models_comparison %>% 
  ggplot() + 
  geom_point(aes(v1.breakpoint, v2.breakpoint))

# There's just one difference. What is it? 
models_comparison %>% 
  mutate(bp_diff = v2.breakpoint - v1.breakpoint) %>% 
  mutate(population = glue::glue("{code}_{x}_{sex}")) %>% 
  ggplot(aes(bp_diff, fct_reorder(population, bp_diff))) + 
    geom_point()

# So, the different starting points make a substantive difference only 
# for FRATNP males, at age 65

# I suspect they some smaller difference for other populations however. Let's 
# check

models_comparison %>% 
  mutate(bp_diff = v2.breakpoint - v1.breakpoint) %>% 
  mutate(population = glue::glue("{code}_{x}_{sex}")) %>% 
  filter(population != "FRATNP_65_male") %>% 
  ggplot(aes(bp_diff, fct_reorder(population, bp_diff))) + 
    geom_point()

# Yes, there are differences, but they are tiny...

@@

I'm now going to compare the breakpoints from the segmented package with the best point from the line/gridsearch approach

```{r}
compare_best_breakmodel_with_null %>% 
  select(code, x, sex, best_tau, all_taus, ftest_pval) %>% 
  unnest(all_taus) %>% 
  ungroup() %>% left_join(
    segmented_breakpoint_models %>% 
      select(code, x, sex, bp_ftest_pval = ftest_pval, bp_breakpoint = breakpoint, bp_breakpoint_se = breakpoint_se)
  ) %>% 
  mutate(population = glue::glue("{code}_{x}_{sex}")) %>% 
  mutate(
    ftest_cat = case_when(
      ftest_pval < 0.01 ~ "<0.01",
      between(ftest_pval, 0.01, 0.05) ~ "0.01-0.05",
      between(ftest_pval, 0.05, 0.10) ~ "0.05-0.10",
      TRUE                            ~ "Not Stat. Sig."
    ),
    ftest_cat = factor(ftest_cat, 
      levels = c("<0.01", "0.01-0.05", "0.05-0.10", "Not Stat. Sig."),
      ordered = TRUE
    )
  ) %>% 
  mutate(
    bp_ftest_cat = case_when(
      bp_ftest_pval < 0.01 ~ "<0.01",
      between(bp_ftest_pval, 0.01, 0.05) ~ "0.01-0.05",
      between(bp_ftest_pval, 0.05, 0.10) ~ "0.05-0.10",
      TRUE                            ~ "Not Stat. Sig."
    ),
    bp_ftest_cat = factor(bp_ftest_cat, 
      levels = c("<0.01", "0.01-0.05", "0.05-0.10", "Not Stat. Sig."),
      ordered = TRUE
    )
  ) %>% 
  ggplot(aes(tau, aic)) +
  geom_line() + 
  geom_point(aes(colour = ftest_cat, alpha = tau == best_tau)) + 
  facet_wrap(~population, scales = "free_y") +
  scale_colour_manual(
    values = c(
      "<0.01" = "darkred", 
      "0.01-0.05"  = "red", 
      "0.05-0.10" = "orange", 
      "Not Stat. Sig." = "lightblue"
      )
  ) + 
  geom_vline(
    aes(
      xintercept = bp_breakpoint,
      colour = bp_ftest_cat  
    )
  ) + 
  geom_rect(
    aes(
      xmin = bp_breakpoint - 1.96 * bp_breakpoint_se, 
      xmax = bp_breakpoint + 1.96 * bp_breakpoint_se
    ),
      ymin = -Inf, ymax = Inf,
    alpha = 0.01
  ) + 
  scale_alpha_manual(values = c(`FALSE` = 0, `TRUE` = 1), guide = "none") +
  labs(
    title = "Penalised fit (AIC) against proposed changepoint year, various population",
    subtitle = "Best fit identified indicate with a point, whose colour indicates statistical significance of changepoint model compared with Null model",
    colour = "Statistical\nsignificance", 
    x = "Year proposed",
    y = "AIC (lower is better)",
    caption = "Vertical lines and bands show estimates and 95% CIs from segmented package"
  )
```

@@

- [X] Jon to figure out why this doesn't work...
    - I think it doesn't work because the `segmented` package functions expect the line to be joined up. Instead it looks like the [structchange](https://cran.r-project.org/web/packages/strucchange/index.html) package can be used for more general model specifications.

@@

```{r, eval = FALSE}
stat_sig_table <- select(tmp, code, x, sex, best_tau, ftest_pval, stat_sig_preference) %>% arrange(ftest_pval)

@@

# Dream Graphs

To finally complete these, without getting distracted/stuck on how to double check and improve upon the underlying data/models

- [ ] I've finally remembered that I can get the breakpoints by aggregating up the changes, as I've done many times previously!

## Graph of ex against year, predicted and observed.

```{r}
# Start with Scotland, female, 0

test_data <- compare_best_breakmodel_with_null %>% 
  filter(code == "GBR_SCO", x == 0, sex == "female") %>% 
  pluck("data", 1)

test_data %>% 
  ggplot(aes(year, ex)) + 
  geom_point()

# Now to get the modelled estimates

compare_best_breakmodel_with_null %>% 
  filter(code == "GBR_SCO", x == 0, sex == "female") %>% 
  pluck("model_alt", 1) %>% 
  broom::tidy()

# So, we want the last year before the breakpoint, then to produce a line that just integrates the (Intercept) term as many years as we want to project forward for.

# So, let's think of a function that makes a graph, and what inputs it needs

show_counterfactual_figure <- function(df, best_tau, alt_model){
  # First we might want to estimate the intercept for the pre-slowdown portion 
  # while fixing the slope at that estimated by (Intercept) in the delta_ex model

  pred_slope <- alt_model %>%  broom::tidy() %>% 
    filter(term == "(Intercept)") %>% 
    pluck("estimate")
  # 
  pred_slope_after <- alt_model %>% 
    broom::tidy() %>% 
    filter(term == "T_paramTRUE") %>% 
    pluck("estimate")

  pred_intercept <-  mean(df$ex - pred_slope * df$year)

  pred_intercept_at_tau <- pred_intercept - pred_slope_after * best_tau

  # I think I should use geom_area for the after part 
  df %>% 
    mutate(
      y_pred_before = pred_intercept + pred_slope * year,
      y_pred_after  = pred_intercept_at_tau + (pred_slope + pred_slope_after) * year
    ) %>% 
    mutate(
      y_pred_before = ifelse(year < best_tau, NA, y_pred_before),
      y_pred_after  = ifelse(year < best_tau, NA, y_pred_after)
    ) %>% 
    ggplot() + 
    geom_point(aes(year, ex)) + 
    geom_vline(xintercept = best_tau) + 
    geom_abline(intercept = pred_intercept, slope = pred_slope) + 
    geom_ribbon(aes(ymax = y_pred_before, ymin = y_pred_after, x = year), alpha = 0.5)
}

this_df <- compare_best_breakmodel_with_null %>% 
  filter(code == "GBR_SCO", x == 0, sex == "male") %>% 
  pluck("data", 1)

this_best_tau <- compare_best_breakmodel_with_null %>% 
  filter(code == "GBR_SCO", x == 0, sex == "male") %>% 
  pluck("best_tau")

this_model_alt <- compare_best_breakmodel_with_null %>% 
  filter(code == "GBR_SCO", x == 0, sex == "male") %>% 
  pluck("model_alt", 1)

show_counterfactual_figure(this_df, this_best_tau, this_model_alt)

# Now to do for all

# Let's also label and set axes intelligently

add_labels_axes <- function(gg, country, start_age, sex, 
                            ylims = list(
                                birth = c(70, 95),
                                x65 = c(0, 25)
                              )
                            ){
  gg + 
    labs(
      x = "Year", 
      y = case_when(
        start_age == 0  ~ "Life expectancy at birth",
        start_age == 65 ~ "Life expectancy at age 65",
        TRUE            ~ NA_character_
      ),
      title = glue::glue("Life expectancy in {country}, at age {start_age}, for {sex}s"),
      subtitle = "Shaded areas show deviation from projected life expectancy before any breakpoint in trend", 
      caption = "Source: Human Mortality Database"
    ) + 
    expand_limits( 
      y = case_when(
        start_age == 0 ~ ylims[["birth"]],
        start_age == 65 ~ ylims[["x65"]],
        TRUE    ~ c(0, 90)
      )
    )
}

compare_best_breakmodel_with_null %>% 
  left_join(country_code_lookup) %>% 
  mutate(
    ggsegplot = pmap(
      .l = list(df = data, best_tau = best_tau, alt_model = model_alt),
      .f = show_counterfactual_figure
    ),
    ggsegplot = pmap(.l = list(gg = ggsegplot, country = country, start_age = x, sex = sex), add_labels_axes)
  ) %>% 
  mutate(NULL = walk(ggsegplot, print))
